\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-28 tex/paper.tex
better toy

\begin{figure*}
    \begin{center}
        \includegraphics[width=0.48\textwidth]{figures/f-plain.pdf}
        \includegraphics[width=0.48\textwidth]{figures/surface-plain.pdf}\\
        \vspace{-0.8cm}
        \includegraphics[width=0.48\textwidth]{figures/f-adversary.pdf}
        \includegraphics[width=0.48\textwidth]{figures/surface-adversary.pdf}
    \end{center}
    \vspace{-1cm}
    \caption{Toy example.
    (Upper left) Conditional probability densities of the decision scores for $Z=0, 1, \dots, 4$
       when $f$ is built without adversarial training. The resulting densities
       are clearly dependent on $Z$, indicating that $f$ is not pivotal.
    (Upper right) The corresponding decision surface, highlighting
       the fact that samples are easier to classify for values of $Z$ closer to $4$,
       hence partially explaining the dependency.
    (Lower left) Conditional probability densities of the decision scores for $Z=0, 1, \dots, 4$ when $f$ is
       built with adversarial training, as outlined in Section~\ref{sec:method}.
       The resulting densities are now almost identical to each other, indicating only a
       small dependency on $Z$.
    (Lower right) The corresponding decision surface, illustrating how adversarial
       training bends the decision function to erase the dependency on $Z$.
    }
    \label{fig:toy}
\end{figure*}

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/training.pdf}
    \vspace{-1cm}
    \caption{Toy example. Training curves for ${\cal L}_f(\theta_f)$, ${\cal L}_r(\theta_f, \theta_r)$
             and ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$.
             Adversarial training was performed for $1000$ iterations, mini-batches of size $M=128$, $K=1$ and $\lambda=10$.}
    \label{fig:toy-training}
\end{figure}

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}(\mu=(0,0), \sigma=2 \times \mathbb{1}) \text{ when } Y=0, \\
    x &\sim {\cal N}(\mu=(1,-1+z), \sigma=2 \times \mathbb{1}) \text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

Assuming a uniform prior over the interval ${\cal Z} = [0;4]$, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we build a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without adversarial
training. As shown in the upper plots of Figure~\ref{fig:toy}, the resulting
classifier is not pivotal, as the conditional probability densities of its
decision scores $f(X;\theta_f)$ show large discrepancies between values $z$ of
the nuisance. While not shown here, a classifier trained only from data
generated at the nominal value $Z=0$ would also not be pivotal.

By contrast, when using Algorithm~\ref{alg:adversarial-training} for training
$f$ jointly with a mixture density network $r$ modeling the conditional
distribution of the nuisance as observed through $f$, the resulting classifier
becomes effectively almost independent on $Z$, as shown in the lower plots of
Figure~\ref{fig:toy}. In particular, the conditional probability densities of
the decision scores $f(X;\theta_f)$ are now very similar to each other,
indicating only a small  dependency on the nuisance, as theoretically expected.
The dynamics of adversarial training is illustrated in
Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$, ${\cal L}_r$ and
${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each iteration of
Algorithm~\ref{alg:adversarial-training}. In the first iterations, we observe
that $f$ is worse than random because it rather tries to maximizes the
adversarial loss ${\cal L}_r$ by producing decision scores $f(X;\theta_f)$ that
are very unlikely under the current model $p_{\theta_r}$, at the expense of
making bad predictions with respect to the classification problem. As learning
goes and $p_{\theta_r}$ becomes more accurate, minimizing $E$ requires making
predictions that are more accurate, hence decreasing ${\cal L}_f$, or that are
less dependent on $Z$, hence shaping $p_{\theta_r}$ so that it becomes more
uniform. Indeed, ${\cal L}_f$ eventually start decreasing, while remaining lower
bounded by $\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed
line in the first plot. Similarly,  ${\cal L}_r$ increasingly tends towards the
differential entropy $H(p_Z) = \log(4)$ of the uniform distribution over ${\cal
Z}$, as shown by the dashed line in the second plot. Finally, let us note that
the ideal situation of a classifier that is both optimal and pivotal appears to
be unreachable for this problem, as shown in third plot by the almost constant
offset between ${\cal L}_f - \lambda {\cal L}_r$ and the dashed line
approximating $\mathbb{E}_{X} [ H(p_{Y|X}) - \lambda H(p_Z)]$.

>>>>

\begin{figure*}
    \begin{center}
        \includegraphics[width=0.48\textwidth]{figures/f-plain.pdf}
        \includegraphics[width=0.48\textwidth]{figures/surface-plain.pdf}\\
        \vspace{-0.8cm}
        \includegraphics[width=0.48\textwidth]{figures/f-adversary.pdf}
        \includegraphics[width=0.48\textwidth]{figures/surface-adversary.pdf}
    \end{center}
    \vspace{-1cm}
    \caption{Toy example.
    (Upper left) Conditional probability densities of the decision scores for $Z=0, 1, 2$
       when $f$ is built without adversarial training. The resulting densities
       are clearly dependent on $Z$, indicating that $f$ is not pivotal.
    (Upper right) The corresponding decision surface, highlighting
       the fact that samples are easier to classify for values of $Z$ closer to $2$,
       hence partially explaining the dependency.
    (Lower left) Conditional probability densities of the decision scores for $Z=0, 1, 2$ when $f$ is
       built with adversarial training, as outlined in Section~\ref{sec:method}.
       The resulting densities are now much closer to each other, indicating only a
       small dependency on $Z$.
    (Lower right) The corresponding decision surface, illustrating how adversarial
       training bends the decision function to erase the dependency on $Z$.
    }
    \label{fig:toy}
\end{figure*}

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/training.pdf}
    \vspace{-1cm}
    \caption{Toy example. Training curves for ${\cal L}_f(\theta_f)$, ${\cal L}_r(\theta_f, \theta_r)$
             and ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$.
             Adversarial training was performed for $300$ iterations, mini-batches of size $M=128$, $K=10$ and $\lambda=10$.}
    \label{fig:toy-training}
\end{figure}

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}(\mu=(0,0), \Sigma= \mathbb{1}) \text{ when } Y=0, \\
    x &\sim {\cal N}(\mu=(1,z), \Sigma= \mathbb{1}) \text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

Assuming a uniform prior over the interval ${\cal Z} = [0;2]$, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we build a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without adversarial
training. As shown in the upper plots of Figure~\ref{fig:toy}, the resulting
classifier is not pivotal, as the conditional probability densities of its
decision scores $f(X;\theta_f)$ show large discrepancies between values $z$ of
the nuisance. While not shown here, a classifier trained only from data
generated at the nominal value $Z=0$ would also not be pivotal.

By contrast, when running Algorithm~\ref{alg:adversarial-training} initialized
from the classifier obtained previously, in combination with a mixture density
network $r$ (modeling $Y|f(X;\theta_f)$ as a truncated mixture of four gaussian
components), adversarial training effectively reshapes the decision function
$f$, so it that becomes almost independent on $Z$, as shown in the lower plots
of Figure~\ref{fig:toy}. In particular, the conditional probability densities of
the decision scores $f(X;\theta_f)$ are now very similar to each other,
indicating only a small  dependency on the nuisance, as theoretically expected.
The dynamics of adversarial training is illustrated in
Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$, ${\cal L}_r$ and
${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each iteration of
Algorithm~\ref{alg:adversarial-training}. In the first iterations, we observe
that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is minimized by
making the classifier less optimal, hence the corresponding increase of ${\cal
L}_f$, but which results in a classifier that is more pivotal, hence the
corresponding increase of ${\cal L}_r$ and the total net benefit. As learning
goes, minimizing $E$ requires making predictions that are more accurate, hence
decreasing ${\cal L}_f$, or that are less dependent on $Z$, hence shaping
$p_{\theta_r}$ so that it becomes more uniform. Indeed, ${\cal L}_f$ eventually
start to slighlty decrease, while remaining lower bounded by $\min_{\theta_f}
{\cal L}_f(\theta_f)$ as approximated by the dashed line in the first plot.
Similarly,  ${\cal L}_r$ keeps tending towards the differential entropy $H(p_Z) =
\log(2)$ of the uniform distribution over ${\cal Z}$, as shown by the dashed
line in the second plot. Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H(p_{Y|X}) -
\lambda H(p_Z)]$.
